In-Band Synchronization for\\Distributed SDN Control Planes

### 2015-09-01 ccr-rev1/inband-sync.tex
bla

The concurrent computing literature offers
a wide range of synchronization abstractions.
E.g., the popular \emph{transactional
  memory} abstraction~\cite{tm-book} provides a collection of
concurrent processes with the ability of aggregating sequences of
shared-memory operations in \emph{atomic  transactions} with
all-or-nothing semantics.

@@

The key idea of our approach is to use the switch configuration space
as a \emph{transactional shared memory}. 
The shared memory of the data-plane configuration space 
can store information about  contention and conflicts
between the controllers.
A transaction can then contain standard OpenFlow control operations
as well as  \emph{synchronization primitives} operating on this shared
memory.
We investigate useful synchronization abstractions and propose 
a simple interface which revolves around policy identifiers (IDs)
which can be claimed and released atomically and hence consistently. 
More generally, our synchronization primitives allow the controllers to define
general notions of conflicts between configuration updates,
not restricted to simple conflicts on the overlapping flow spaces.
In particular, we can define dependencies between mappings of
independent flows to the physical infrastructure, which is important,
e.g., for load-balancing applications.

We complement our conceptual contributions by
demonstrating that our approach can readily be implemented
in today's standard \emph{OpenFlow}
 protocol (version 1.4~\cite{openflow}).
%
%By providing more complex and atomic transactions over the
%switch configuration
%space, several synchronization problems can be resolved.
%This paper investigates the implementation of a transactional interface
%(\`{a} la STN~\cite{stn}) modified concurrently by multiple controllers.
%\petr{"interface modified concurrently" does not sound right -> "accessed concurrently"?}
%
In particular, we show that OpenFlow's match-action paradigm can be instrumented
to realize simple key-value stores in the configuration space.

@@

Using our abstraction, the controllers can
solve several important synchronization problems.
We describe an OpenFlow implementation of the fundamental
\emph{compare-and-set} (CAS) primitive,
which in turn can be used to solve consensus and
implement a generic replicated control plane service~\cite{Her91} in a consistent and
fault-tolerant way: a mainstream building block in
modern system design.
We also discuss a simple CAS-based concurrent policy-update mechanism,
and use our synchronization abstraction to provide
the missing link for the read-modify-write object
postulated in~\cite{stn}. %\petr{We do not seem to discuss the latter?}

@@

\noindent\textbf{Organization}
The rest of the paper is organized as follows.
In Section~\ref{sec:background}, we review the basics of the  OpenFlow
protocol and discuss the features which are relevant for our work.
In Section~\ref{sec:motivation}, we highlight the limitations of the OpenFlow
primitives and motivate the need for transactions.
We present our approach in detail in Section~\ref{sec:main}, and
discuss use cases in Section~\ref{sec:apps}.
%Section~\ref{sec:compo} describes how to perform in-band composition.
We
%review related literature in Section~\ref{sec:relwork}, and
conclude our paper
in Section~\ref{sec:conclusion}.
% Some technical details are postponed
%to the Appendix.

@@

\section{Background: SDN and OpenFlow}\label{sec:background}
%\petr{ the 1st sentence does not belong here}\liron{I think it serves as a motivation for the section, but it fines by me to remove it}

@@

\section{Limitations and Motivation}\label{sec:motivation}

At first sight, bundling \emph{FlowMod} commands with  the $\checko$
flag already provides a useful synchronization mechanism.
Indeed, its ``mini-transaction'' nature allows a controller
to install multiple flow entries in an atomic way.
%[[PK we did not say what reads and writes are...
%At first sight, the OpenFlow bundle feature seems to provide everything
%we need to implement our transactional interface \petr{did we say what we need already?}: the bundle
% defines a sequence of read and write operations
% on the switch state which is executed atomically.
%]]
However, as is,
%while the bundle can indeed be seen as a ``mini-transaction'',
bundling has important limitations, as we will argue in the following.
%\petr{unclear what "reads and writes" are here, not discussed before. How openflow can be used to implement reads and writes?}

First, %if employed without any additional features,
except for some limited constraint types (such as
overlapping flow spaces or insufficient free space),
a bundle does not provide any means to modify the switch configuration
based on application-specific conditions, which depend on the
current configuration,
However, as we will argue, support for more generic conditions under which configuration
updates
should or should not take place, is often desirable.
%
%But also bundling FlowMod commands with the check overlap feature ($\checko$)
%comes with limitations: while it is possible
% to \emph{abort} a bundle depending on certain conditions
% in the \emph{overlapping
% flow spaces}, oftentimes it is desirable to be able
% to define more general conditions and notions of conflicts
% under which an atomic update should not take effect.
% \liron{replaced transaction in the last sentence with " atomic update"}
%
Second, sometimes it is desirable to react to conflicts in smarter ways than
by simply aborting an operation, e.g., by supporting \emph{conditional modifications}.
Let us consider two examples.

%\begin{enumerate}
\vspace{1mm}
\noindent \textbf{Example~1: Handling complex dependencies.} Let us consider two controllers
in charge
of \emph{load-balancing}, see Figure~\ref{fig:CAS-example}
for a simplistic example with two links.
In such a scenario, seemingly independent actions, defined over
completely independent logical flow spaces (say, two different
TCP micro-flows),
%originated in disjoint ranges of IP addresses),
may actually be dependent: the flows share the underlying physical network.
Accordingly, if multiple controllers concurrently and independently
update forwarding rules according to a na\"ive load-balancing algorithm \`{a} la Algorithm~\ref{alg:naive-lb},
they may involuntarily
unbalance the flow allocation.
Simple flow space overlap checks cannot be used to detect
such conflicts.
Note that, assuming that all these flows are
completely independent, bundling and $\checko$ features do not really
help here.

{\small
\begin{algorithm}[t]
    \caption{$\textit{Na\"ive Flow Balancing}$}
    \label{alg:naive-lb}
\begin{algorithmic}[1]

    \Require \emph{new\_flows}: flow entries of the new flows,
    \emph{ports}: a list of switch output ports,
    \State $\textit{policy} \gets \textbf{read\_config}$
    \State $\textit{update\_cmds} \gets \textsc{lb\_update}(policy)$ %(new\_flows, ports, policy)$
    \State $\textbf{send}\{\textit{update\_cmds}\}$
%    \State $\textbf{execute-atomic}\{update\_cmds\}$
    \Statex

    \Statex \emph{\bf Function} $\textsc{lb\_update}(\textit{policy})$ % (new\_flows, ports, policy)}$

	\State $\textit{flows\_per\_port} \gets [0, 0, \ldots, 0]$
	\ForAll {$\textit{entry} \in \textit{policy}$}
	    \State $\textit{port} \gets \textit{entry.output\_port}$
	    \State $\textit{flows\_per\_port}[\textit{port}]++$
	\EndFor
	\State $\textit{new\_flow\_mod\_cmds} \gets \{\}$
	\For {$\textit{entry} \in \textit{new\_flows}$}
	    \State $\textit{port} \gets \arg\min_{p\in \textit{ports}} \textit{flows\_per\_port}[p]$
	    \State $\textit{entry.output\_port} \gets \textit{port}$
	    \State $\textit{flows\_per\_port}[\textit{port}]++$
	    \State $\textit{cmd} \gets \textit{FlowMod}(add,f)$
	    \State $\textit{new\_flow\_mod\_cmds}.\textit{add}(\textit{cmd})$
	\EndFor
	\Return $\textit{new\_flow\_mod\_cmds}$
  %  \State \emph{\bf EndFunction}
\end{algorithmic}
\end{algorithm}
}

\vspace{1mm}
\noindent\textbf{Example~2: Supporting modifications.}
Consider the arguably very basic problem of \emph{modifying} previously installed forwarding rules
in a consistent manner.
In principle, a bundle allows us to conditionally add new rules or remove
old rules, and, using the flow space overlap
check feature $\checko$, to abort the transaction if potentially conflicting
rules were already there. However, it is not obvious how
to use the bundle to
modify \emph{existing} rules, especially if the modification depends on
specific conditions.
In other words, the bundle and overlap check only help
for the \emph{conditional creation} of rules, not their \emph{modification}.

\vspace{2mm}
\noindent\textbf{Our Approach.}
This paper proposes mechanisms to solve these two examples,
by using a minimal
``meta-configuration'' space. That is, controllers store information
on contention and conflicts between
concurrently installed policy updates on the switch. Controllers can read and modify
this meta-data by applying synchronization primitives.
In the following section, we
describe how this space can be organized and consistently maintained
by multiple controllers despite of concurrency and conflicts.

We will later come back to these examples, and show that Algorithm~\ref{alg:naive-lb}
can actually be used without any modifications within our framework.
Consistent load-balancing can also be achieved using our atomic Compare-and-Set instruction based on policy ids (pids),
see Figure~\ref{fig:CAS-example} (\emph{right}).

\section{Synchronization Abstractions}\label{sec:main}

We now describe how our  meta-configuration space at
an OpenFlow switch can be implemented and manipulated using our
synchronization framework.

The meta-configuration is implemented as a set of flow entries
designed in a way that it does not affect the processing of data plane
traffic. One way to achieve this is to set a low priority to
the meta flow entries and ensure that the regular flow entries
maintain higher priority levels.
Alternatively, we could leverage the possibility to maintain multiple
flow tables at an OpenFlow switch, and use a separate flow table for
the meta-configuration,  making sure that the
\emph{pipelining} mechanism will never involve this
flow table.

%\petr{How to make sure that traffic is not affected by the ``meta'' part?}
\vspace{1mm}
\noindent\textbf{Synchronization.}
 %
Our framework provides the controllers with the ability to execute
sequences of \emph{operations} on the
switch configuration in an \emph{atomic} (transactional) or \emph{non-atomic} mode.
In the atomic mode, either \emph{all} operations in the sequence, henceforth simply
called \emph{transaction},
should be executed, or none (the transaction is rejected).
In the non-atomic mode, no atomicity guarantees are provided.
An operation here can be a regular FlowMod command or a synchronization
primitive, which, as we
show below, translates into a sequence of FlowMod commands operating on the
switch meta-configuration.

@@

This paper initiated the study of mechanisms which allow distributed controllers
to synchronize their possibly concurrent actions. We believe that our in-band
approach is natural and attractive as it can avoid the
complexities and limitations of distributed out-of-band agreement protocols.
Moreover, our mechanisms are simple and can be implemented using the \emph{standard OpenFlow}
protocol: they do not require any protocol/hardware extensions as postulated in recent literature~\cite{stn,netpaxos}.
Our work may hence also contribute to the ongoing discussion of what can be implemented
in-band in today's OpenFlow protocol~\cite{reclaim},
as well as to useful high-level concurrency objects and
language abstractions.~\cite{pyretic}
We also plan to release a small library with our synchronization primitives. 
Moreover, we are investigating possible generalizations of our transactional guarantees,
across multiple switches~\cite{stn}, as well as opportunities for more fine-grained
\emph{partial} updates that concern subsets of flow entries, thus improving concurrency.

>>>>

The concurrent computing literature offers
a wide range of synchronization abstractions
to realize consistent distributed systems.
In particular, the popular \emph{transactional
  memory} abstraction~\cite{tm-book} provides a collection of
concurrent processes with the ability of aggregating sequences of
shared-memory operations in \emph{atomic  transactions} with
all-or-nothing semantics.

@@

The key idea of our approach is to use the switch configuration space
as a \emph{transactional shared memory}. 
The shared memory of the data-plane configuration space 
can store information about  contention and conflicts
between the controllers.
A transaction can then contain standard control and update operations
as well as  \emph{synchronization primitives} operating on this shared
memory.
We investigate useful synchronization abstractions and propose 
a simple interface which revolves around policy identifiers (IDs)
which can be claimed and released atomically and hence consistently. 
More generally, our synchronization primitives allow the controllers to define
general notions of conflicts between configuration updates,
not restricted to simple conflicts on the overlapping flow spaces.
In particular, we can define dependencies between mappings of
independent flows to the physical infrastructure, which is important,
e.g., for load-balancing applications.

%
%By providing more complex and atomic transactions over the
%switch configuration
%space, several synchronization problems can be resolved.
%This paper investigates the implementation of a transactional interface
%(\`{a} la STN~\cite{stn}) modified concurrently by multiple controllers.
%\petr{"interface modified concurrently" does not sound right -> "accessed concurrently"?}
%

@@

As a case study, and to demostrate the feasibility
of our conceptual approach, we 
consider the standard \emph{OpenFlow}
 protocol (version 1.4~\cite{openflow}).
 In particular, we show that OpenFlow's match-action paradigm can be instrumented
to realize simple key-value stores in the configuration space.

Using our abstraction, the controllers can
solve several important synchronization problems.
For example, we describe an OpenFlow implementation of the fundamental
\emph{compare-and-set} (CAS) primitive.
The atomic CAS operation is known to provide an infinite consensus number,
and can for instance be used to 
implement generic replicated control plane services~\cite{Her91} in a consistent and
fault-tolerant way: a mainstream building block in
modern distributed system design (Chubby, Google Spanner, Amazon Webservices, etc.).
We discuss a simple CAS-based concurrent policy-update mechanism,
and present a generic template to implement concurrent policy
compositions of a user-specic update functions.
Our synchronization abstraction also provides
the missing link for the read-modify-write object
postulated in~\cite{stn}. %\petr{We do not seem to discuss the latter?}

@@

\noindent\textbf{Organization}
The rest of the paper is organized as follows.
In Section~\ref{sec:motivation}, we motivate the 
transactional approach.
We present our abstractions in more detail in Section~\ref{sec:main}.
In Section~\ref{sec:background}, we discuss
an OpenFlow implementation, 
and we 
discuss use cases in Section~\ref{sec:apps}.
%Section~\ref{sec:compo} describes how to perform in-band composition.
We
%review related literature in Section~\ref{sec:relwork}, and
conclude our paper
in Section~\ref{sec:conclusion}.
% Some technical details are postponed
%to the Appendix.

@@

\section{Motivation and Example}\label{sec:motivation}

A transactional distributed system offers all-or-nothing semantics, 
and aborts transactions in case of conflicts. 
For instance, in the context of SDNs, a transaction trying to add a forwarding rule whose
domain overlaps with an existing rule of the same priority
can constitute a conflict. 
However, as we will argue, there are often conflicts which are less obvious,
and go beyond simple range overlaps. 
Moreover, sometimes it is desirable to react to conflicts in smarter ways than
by simply aborting an operation, e.g., by supporting \emph{conditional modifications}.
Let us consider two examples.

%
%But also bundling FlowMod commands with the check overlap feature ($\checko$)
%comes with limitations: while it is possible
% to \emph{abort} a bundle depending on certain conditions
% in the \emph{overlapping
% flow spaces}, oftentimes it is desirable to be able
% to define more general conditions and notions of conflicts
% under which an atomic update should not take effect.
% \liron{replaced transaction in the last sentence with " atomic update"}
%

%Let us consider an example.

%\begin{enumerate}
\vspace{1mm}
\noindent \textbf{Example: Handling complex dependencies.} We consider two controllers
in charge
of \emph{load-balancing}, see Figure~\ref{fig:CAS-example}
for a simplistic example with two links.
In such a scenario, seemingly independent actions, defined over
completely independent logical flow spaces (say, two different
TCP micro-flows),
%originated in disjoint ranges of IP addresses),
may actually be dependent: the flows share the underlying physical network.
Accordingly, if multiple controllers concurrently and independently
update forwarding rules according to a na\"ive load-balancing algorithm,
they may involuntarily
unbalance the flow allocation.
Simple flow space overlap checks cannot be used to detect
such conflicts.

This paper proposes mechanisms to solve these synchronization problems 
by using a minimal
``meta-configuration'' space. That is, controllers store information
on contention and conflicts between
concurrently installed policy updates on the switch. Controllers can read and modify
this meta-data by applying synchronization primitives.
In the following section, we
describe how this space can be organized and consistently maintained
by multiple controllers despite of concurrency and conflicts.

\section{In-Band Synchronization}\label{sec:main}

In order to synchronize controllers, we propose
to use a meta-configuration space at
the switches which can be manipulated using our
synchronization framework.

The meta-configuration is implemented as a set of flow entries
designed in a way that it does not affect the processing of data plane
traffic. One way to achieve this, for example, is to set a low priority to
the meta flow entries and ensure that the regular flow entries
maintain higher priority levels. We will discuss other options later
in the context of OpenFlow.

%\petr{How to make sure that traffic is not affected by the ``meta'' part?}
\vspace{1mm}
\noindent\textbf{Synchronization.}
 %
We provide the controllers with the ability to execute
sequences of \emph{operations} on the
switch configuration in an \emph{atomic} (transactional) or \emph{non-atomic} mode.
In the atomic mode, either \emph{all} operations in the sequence, henceforth simply
called \emph{transaction},
should be executed, or none (the transaction is rejected).
In the non-atomic mode, no atomicity guarantees are provided.

\section{Case Study: OpenFlow}\label{sec:background}
%\petr{ the 1st sentence does not belong here}\liron{I think it serves as a motivation for the section, but it fines by me to remove it}

@@

@@

This paper initiated the study of mechanisms which allow distributed controllers
to synchronize their possibly concurrent actions. We believe that our centralized in-band
approach is natural and attractive, as it avoids the
complexities and limitations of distributed out-of-band agreement protocols, which may also
introduce additional points of potential failures.
Moreover, our mechanisms are simple and can be implemented using the \emph{standard OpenFlow}
protocol: they do not require any protocol/hardware extensions as postulated in recent literature~\cite{stn,netpaxos}.
Our work may hence also contribute to the ongoing discussion of what can be implemented
in-band in today's OpenFlow protocol~\cite{reclaim},
as well as to useful high-level concurrency objects and
language abstractions.~\cite{pyretic}
In general, we understand our paper as a first step, and more
research is required to investigate and compare useful
inband synchronization abstractions, also in terms of performance.
%In the long run, we seek to develop in-band mechanisms
%which allow us to seemlessly transition between single controller and
%multiple controller mode.

%We also plan to release a small library with our synchronization primitives. 
%Moreover, we are investigating possible generalizations of our transactional guarantees,
%across multiple switches~\cite{stn}, as well as opportunities for more fine-grained
%\emph{partial} updates that concern subsets of flow entries, thus improving concurrency.
